---
title: "Reproducible Research: Peer Assessment 1"
author: Suzannah Herron
output: 
  html_document:
    keep_md: true
---

## Data Overview

*Activity Monitoring Dataset* with:  
- **steps**: Number of steps taken in a 5-min interval (missing values coded as `NA`)  
- **date**: Date in YYYY-MM-DD format\
- **interval**: Identifier for the 5-min interval in which measurement was taken

## 1. Loading and preprocessing the data

```{r loading, message=FALSE}
file_name <- "activity.zip"
if (!file.exists(file_name)) {
    url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
    download.file(url, file_name, method = "curl")  
}

if (!file.exists("activity.csv")) {
    unzip(file_name)
}

activity <- read.csv("activity.csv")
```
Change `date` column from characters strings.
```{r message=TRUE}
library(lubridate)
activity$date <- ymd(activity$date)
summary(activity)
```
There are 2304 `NA` values

## 2. What is mean total number of steps taken per day?

Removing NA values
```{r}
act_na <- activity[!is.na(activity$steps), ]
summary(act_na)
```
\newpage

### 2.1 Calculate total number of steps taken per day

```{r}
act_sum <- aggregate(steps ~ date, act_na, sum)
head(act_sum)
```
```{r message=FALSE}
library(ggplot2)
```

### 2.2 Graph of Sum of Steps Taken Per Day


```{r}
png(filename = "total_steps_hist.png", width = 480, height = 480, bg = NA)
ggplot(data = act_sum) +
  geom_histogram(mapping = aes(steps), fill = "violetred", bins = 7, color = "black") +
  labs(title = "Total Steps Taken Per Day",
       x = "Total Steps",
       y = "Number of Days")
dev.off()
```
\newpage

### 2.3 Calculate average number of steps taken per day

```{r}
act_mean <- mean(act_sum$steps)
act_med <- median(act_sum$steps)
act_mean
act_med
```
The mean of total steps per day is `r act_mean` and the median is `r act_med`.

\newpage

## 3. What is the average daily activity pattern?

### 3.1 Make a time series plot of 5-min interval and avg num of steps taken averaged across all days

Calculate mean of steps across intervals
```{r}
int_avg <- aggregate(steps ~ interval, act_na, mean)
head(int_avg)
```
```{r}
png(filename = "avg_intervals.png", width = 480, height = 480, bg = NA)
ggplot(data = int_avg) +
  geom_line(mapping = aes(x = interval, y = steps), color = "purple") +
  labs(title = "Avg. Steps Per 5-Min Interval",
       x = "5-Min Interval",
       y = "Avg. Steps")
dev.off()
```

### 3.2 Which 5-min interval has most steps?

```{r}
int_max <- int_avg[which.max(int_avg$steps), 1]
int_max 
```
The 5-minute interval with the most steps is `r int_max`.


## 4. Imputing missing values

### 4.1 Calculate the total number of `NA`s

```{r}
sum(is.na(activity))
```
The total number of `NA`s in the dataset is 2304.\
Alternatively, `summary()` will give you this info as well:
```{r}
summary(activity)
```


### 4.2 Devise strategy for filling in missing values and create new dataset

Find mean of average of steps per interval.
```{r}
avg <- mean(int_avg$steps)
avg
```
Create df with the `NA`s.
```{r}
nas <- is.na(activity$steps)
```

Insert `avg` into values with `NA`
```{r}
data_avg <- activity
data_avg[nas, 1] <- avg
head(data_avg)
```
Use new `data_avg` dataset to calculate sum of step per day
```{r}
data_sum <- aggregate(steps ~ date, data_avg, sum)
head(data_sum)
```




### 4.3 Make histogram of total number of steps taken each day. Report mean and median. How does it differ?

```{r}
png(filename = "total_per_day.png", width = 480, height = 480, bg = NA)
ggplot(data = data_sum) +
  geom_histogram(mapping = aes(steps), fill = "black", bins = 7, color = "violetred") +
  labs(title = "Total Steps Taken Per Day",
       x = "Total Steps",
       y = "Number of Days")
dev.off()
```

```{r}
data_mean <- mean(data_sum$steps)
data_med <- median(data_sum$steps)
data_mean
data_med
```
The mean without the `NA` values is `r act_mean` and with inserted averages, it is `r data_mean`. \
The median without the `NA` values is `r act_med` and with inserted averages, it is `r data_med`.

```{r}
diff_mean <- act_mean - data_mean
diff_med <- act_med - data_med
diff_mean
diff_med
```
There is no impact extracting the `NA`s to the mean and a very negligible difference of `r diff_med` to the median.

## 5. Are there differences in activity patterns between weekdays and weekends?

### 5.1 Using `data_sum`, create a new variable factor with two levels of "weekday" and "weekend".



```{r message=FALSE}
library(dplyr)
```
Using `weekdays()` to find day of the week then changing "Saturday" and "Sunday" to "Weekend" and the rest to "Weekday".
```{r}
weekend <- mutate(data_avg, weekday = ifelse((weekdays(data_avg$date) == "Saturday" | weekdays(data_avg$date) == "Sunday"), "Weekend", "Weekday"))
head(weekend)
```
### 5.2 Make a plot of weekends vs. weekdays

```{r}
png(filename = "weekend_intervals.png", width = 480, height = 480, bg = NA)
ggplot(data = weekend) +
  geom_line(mapping = aes(x = interval, y = steps), color = "purple") +
  facet_wrap(~weekday) +
  labs(title = "Weekday vs. Weekend",
       x = "5-Min Interval",
       y = "Avg. Steps")
dev.off()
```












